Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-03-22 VegOzone.tex
Spell-checked

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects on the model predicted ozone uptake with respect to the MM default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is in any case larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studyed here. Although our temperatue acclimations are hypothetical, this indicates a possibly considerable underestimation of ozone risk for subarctic species. Especially, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the ozone damage risk for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displayes a higher ozone damage risk than 2019. However, taking the later onset and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this ia a bug or a feature of the $\mathrm{DO_3SE}$ model is not yet clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold with respect to MM ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The estimated biomass reduction in \unit{\%} for the bespoke temperature response function of type boreal in accordance to \citet{ICP:MappingManual2017, ESPR:Hayes2021} is shown in Table~\ref{tab:biomass_reduction}. The estimated uncertainty range corresponds to the deviation given by the $\pm 20\,\unit{\%}$ variations of the light threshold and SWP=on. For downy birch and perennial grassland, substantial reductions in total biomass are predicted for any of the two years. Norway spruce is least affected. The systematic uncertainties for downy birch are one order of magnitude larger than for Norway spruce and perennial grassland. For all species or PFTs the interannual variability is larger than the systematic uncertainties estimated for a specific acclimation and is of the same order of magnitude as the divergance between the hypothetical acclimations. For comparison, we averaged the corresponding biomass reduction for the MM temperature and light response functions with both SWP on/off but with bespoke GS over both years, we find a non negligible, which means above systematic uncertainties, low-bias for both downy birch and perennial grassland with respect to the MM default parameteraziation. For Norway spruce, the default MM parameterization average lies within the given systematic uncertainty range, this may reflect the more comprehensive research on this species or PFT.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for bespoke temperature response function of type boreal with bespoke GS and SWP=off. The uncertainty ranges reported here correspond to maximum a divergance by varying the light threshold and turning SWP sensitivity on in $\mathrm{DO_3SE}$. For comparison, the corresponding biomass reduction for the MM temperature and light response functions, SWP on/off, and bespoke GS averaged over both years is shown with its standard deviation.}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Downy birch & Norway spruce & \multicolumn{2}{c}{Perennial grassland}\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $^*12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $^*13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $^*9.6\pm 1.4$\\
\bottomhline
\end{tabular}
\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}

\conclusions[Discussion and conclusions]
\label{sec:conc}
We have examplarily studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. We also assessed implications on potential ozone damage risks arising at the advant of climate change. We found that conditions for ozone formation were more favourable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occured more frequently and at higher levels in 2018. This was in particular caused by the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).

We found that the length and timing of the growing season is crucial for risk assessment based on accumulated ozone flux metrices ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for Norway spruce in 2018 to be at least one month shorter than in 2019. The found onsets ($106/122\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the onset of growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods to estimators show a later onset for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

We found representative ozone climatologies (multi-annual means) for northern Fennoscandia and Svanhovd based on surface ozone monitoring. To account for missing data from our 2018 record, we proposed a routine for gap filling of ozone observations following a Reynolds's decomposition into annomalies and climatology. We also assessed the quality of available global reanalysis (ECWMF MACC and CAM) and regional model reanalysis (Copernicus) near-surface ozone concentrations products and conclude that only the Copernicus product can sufficiently reproduce observed ozone concentrations in northern Fennoscandia.

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were not significantly elevated by statistical means. It therefor occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrices like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periodes of time. Consequently, accute cellular damage caused by very high fluxes over a short periode of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differenciated picture. For herbal and grassland species (similar for crops) a strong dependency on phenological state, e.g. high susceptibility to ozone in the reproductive state, were found \citep{EP:Bassin2004}.

To estimate potential biases in assessment of ozone induced risk on Fennoscandic vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandic vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation to increase the enclosed area between the respective response functions and the PDF. These hypothetical species or PFTs still need varification by experiemental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Only accounting for acclimation, the predicted ozone uptake is typically larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland. However, taking bespoke length and timing of growing season into account the opposite might be the case under certain conditions, e.g. a later start of the growing season with respect to the ozone spring peak for downy birch annd perennial grassland. Simulations of Norway spruce with bespoke (shorter than default) growing season show only little sensitivity to acclimations and inter-annual variation of meteorological conditions. We have not assessed the effect of VPD response function here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

The impact of the timing of the growing season can be of the same order of magnitude as the acclimation of response functions but points into the opposite direction. Especially, the trend towards an earlier onset of the growing season in northern Fennoscandia with respect to the ozone spring peak induces large uncertainties for future risk assessment and includes potential higher ozone damage risk. However, the decline of ozone concentrations in arctic spring is partly caused by the removal of ozone through active vegetation at the start ot the growing season. It is therefor not clear if the peak concentrations of ozone experienced by vegetation in the future with an earlier onset of the growing season, will be reduced compared to present day. Hence, a simulated earlier onset of growing season with present day ozone concentration might overestimate ozone uptake and hence ozone induced risk of damage.

Based on the methodes described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For Norway spruce, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care, because the biomass reduction functions have been evaluated based on central European species or PFTs and might as well be subject to biases.

At the same time, intra-specied variability even at the same location is typically non neglicable \citep{EP:Bassin2004,Amb:Girgzdiene2009}. This may explain contradictory results regarding the ozone sensitivity of natural vegetation in northern Europe. While \citet{FS:Subramanian2014} reported a modeled reduction of net primary production in Norway spruce ($4.3-15.5\,\unit{\%}$) and birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, Lithuanian forest decidous trees were observed to be more susceptible to visible damage by ozone than Norway spruce \citep{Amb:Girgzdiene2009}. Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed to elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one-season exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.
In our study, we do not explicitly account for this natural distribution of individual plant traits, but the chosen hypothetical, bespoke response functions might reflect this diversity within a given species or PFT. The resiliance of northern Fennoscandic plant communities to climate change and predictions in this regard strongly depends on both the existing range of acclimations as well as the actual type of acclimation.

Cumulative ozone uptake has been reported to reduce stomatal conductance significatly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affacted. This might be due competing effects, as an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the $\mathrm{DO_3SE}$ model does not account for any ozone inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.

Beyond the risk assessment of ozone induced damage, our hypothetical bespoke temperature reponse function may have important indications for landsurface modeling, where problems in productivity of species especially in the Arctic regions occure due to PFTs which are not suitable for the climate. We pointed out with respect to default perennial grassland parameterization that that did not show any substential stomatal conductance in 2019 which we characterized as a normal year. In termas of GPP this would mean unrealistically low carbon sequestation. An automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

@@

\dataavailability{All data is available from public databases or through institutional access. $\mathrm{DO_3SE}$ modeling results can be made available upon request {\bf TODO: Upload to an archieve together with processed, gapfilled input data?}.} %% use this section when having only data sets available

@@

\authorcontribution{SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV helped in framing the research articel, has performed on-site observation of vegetation damage induced by ozone, advised with regard to plant physiological processes and contributed text to the introduction. LE has contributed to discussions, setup and evaluation of $\mathrm{DO_3SE}$ modeling results, and development of the PDF-based temperature acclimation. CO has collected species or PFT parameters from literature, performed all $\mathrm{DO_3SE}$ simulations and initial graphical checks of the results. ALEB contributed with her experience regarding subartic vegetation in Finnmark. FS contributed with a summary of the 2018 meteorological conditions. TB gave guidance in a broader research sense. All authors contributed to discussions, writing, and conclusions.} %% this section is mandatory

>>>>

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects on the model predicted ozone uptake with respect to the MM default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is in any case larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studied here. Although our temperature acclimations are hypothetical, this indicates a possibly considerable underestimation of ozone risk for subarctic species. Especially, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the ozone damage risk for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

For downy birch (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Downy birch displays the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the boreal type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later onset and overall shorter GS in 2018 into account, this is reversed for the boreal acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not yet clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The overall bias deduced from bespoke GS for boreal and cold with respect to MM ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019 with maximum systematic uncertainties due to the variation of the light threshold of $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

@@

The estimated biomass reduction in \unit{\%} for the bespoke temperature response function of type boreal in accordance to \citet{ICP:MappingManual2017, ESPR:Hayes2021} is shown in Table~\ref{tab:biomass_reduction}. The estimated uncertainty range corresponds to the deviation given by the $\pm 20\,\unit{\%}$ variations of the light threshold and SWP=on. For downy birch and perennial grassland, substantial reductions in total biomass are predicted for any of the two years. Norway spruce is least affected. The systematic uncertainties for downy birch are one order of magnitude larger than for Norway spruce and perennial grassland. For all species or PFTs the interannual variability is larger than the systematic uncertainties estimated for a specific acclimation and is of the same order of magnitude as the divergence between the hypothetical acclimations. For comparison, we averaged the corresponding biomass reduction for the MM temperature and light response functions with both SWP on/off but with bespoke GS over both years, we find a non negligible, which means above systematic uncertainties, low-bias for both downy birch and perennial grassland with respect to the MM default parameterization. For Norway spruce, the default MM parameterization average lies within the given systematic uncertainty range, this may reflect the more comprehensive research on this species or PFT.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for bespoke temperature response function of type boreal with bespoke GS and SWP=off. The uncertainty ranges reported here correspond to maximum a divergence by varying the light threshold and turning SWP sensitivity on in $\mathrm{DO_3SE}$. For comparison, the corresponding biomass reduction for the MM temperature and light response functions, SWP on/off, and bespoke GS averaged over both years is shown with its standard deviation.}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Downy birch & Norway spruce & \multicolumn{2}{c}{Perennial grassland}\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $^*12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $^*13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $^*9.6\pm 1.4$\\
\bottomhline
\end{tabular}
\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}

\conclusions[Discussion and conclusions]
\label{sec:conc}
We have exemplarily studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. We also assessed implications on potential ozone damage risks arising at the advent of climate change. We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was in particular caused by the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).

We found that the length and timing of the growing season is crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for Norway spruce in 2018 to be at least one month shorter than in 2019. The found onsets ($106/122\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the onset of growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods to estimators show a later onset for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

We found representative ozone climatologies (multi-annual means) for northern Fennoscandia and Svanhovd based on surface ozone monitoring. To account for missing data from our 2018 record, we proposed a routine for gap filling of ozone observations following a Reynolds's decomposition into anomalies and climatology. We also assessed the quality of available global reanalysis (ECMWF MACC and CAM) and regional model reanalysis (Copernicus) near-surface ozone concentrations products and conclude that only the Copernicus product can sufficiently reproduce observed ozone concentrations in northern Fennoscandia.

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were not significantly elevated by statistical means. It therefor occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrics like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periods of time. Consequently, acute cellular damage caused by very high fluxes over a short period of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differentiate picture. For herbal and grassland species (similar for crops) a strong dependency on phenological state, e.g. high susceptibility to ozone in the reproductive state, were found \citep{EP:Bassin2004}.

To estimate potential biases in assessment of ozone induced risk on Fennoscandic vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandic vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation to increase the enclosed area between the respective response functions and the PDF. These hypothetical species or PFTs still need verification by experimental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Only accounting for acclimation, the predicted ozone uptake is typically larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland. However, taking bespoke length and timing of growing season into account the opposite might be the case under certain conditions, e.g. a later start of the growing season with respect to the ozone spring peak for downy birch and perennial grassland. Simulations of Norway spruce with bespoke (shorter than default) growing season show only little sensitivity to acclimations and inter-annual variation of meteorological conditions. We have not assessed the effect of VPD response function here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

The impact of the timing of the growing season can be of the same order of magnitude as the acclimation of response functions but points into the opposite direction. Especially, the trend towards an earlier onset of the growing season in northern Fennoscandia with respect to the ozone spring peak induces large uncertainties for future risk assessment and includes potential higher ozone damage risk. However, the decline of ozone concentrations in arctic spring is partly caused by the removal of ozone through active vegetation at the start of the growing season. It is therefor not clear if the peak concentrations of ozone experienced by vegetation in the future with an earlier onset of the growing season, will be reduced compared to present day. Hence, a simulated earlier onset of growing season with present day ozone concentration might overestimate ozone uptake and hence ozone induced risk of damage.

Based on the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For Norway spruce, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care, because the biomass reduction functions have been evaluated based on central European species or PFTs and might as well be subject to biases.

At the same time, intra-species variability even at the same location is typically non negligible \citep{EP:Bassin2004,Amb:Girgzdiene2009}. This may explain contradictory results regarding the ozone sensitivity of natural vegetation in northern Europe. While \citet{FS:Subramanian2014} reported a modeled reduction of net primary production in Norway spruce ($4.3-15.5\,\unit{\%}$) and birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, Lithuanian forest decidous trees were observed to be more susceptible to visible damage by ozone than Norway spruce \citep{Amb:Girgzdiene2009}. Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed to elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one-season exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.
In our study, we do not explicitly account for this natural distribution of individual plant traits, but the chosen hypothetical, bespoke response functions might reflect this diversity within a given species or PFT. The resilience of northern Fennoscandic plant communities to climate change and predictions in this regard strongly depends on both the existing range of acclimations as well as the actual type of acclimation.

Cumulative ozone uptake has been reported to reduce stomatal conductance significantly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affected. This might be due competing effects, as an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the $\mathrm{DO_3SE}$ model does not account for any ozone inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.

Beyond the risk assessment of ozone induced damage, our hypothetical bespoke temperature response function may have important indications for land-surface modeling, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out with respect to default perennial grassland parameterization that that did not show any substantial stomatal conductance in 2019 which we characterized as a normal year. In terms of GPP this would mean unrealistically low carbon sequestration. An automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

@@

\dataavailability{All data is available from public databases or through institutional access. $\mathrm{DO_3SE}$ modeling results can be made available upon request {\bf TODO: Upload to an achieve together with processed, gap filled input data?}.} %% use this section when having only data sets available

@@

\authorcontribution{SF has written the manuscript, collected and processed ozone and environmental data, and performed all statistical analyses. AVV helped in framing the research article, has performed on-site observation of vegetation damage induced by ozone, advised with regard to plant physiological processes and contributed text to the introduction. LE has contributed to discussions, setup and evaluation of $\mathrm{DO_3SE}$ modeling results, and development of the PDF-based temperature acclimation. CO has collected species or PFT parameters from literature, performed all $\mathrm{DO_3SE}$ simulations and initial graphical checks of the results. ALEB contributed with her experience regarding subarctic vegetation in Finnmark. FS contributed with a summary of the 2018 meteorological conditions. TB gave guidance in a broader research sense. All authors contributed to discussions, writing, and conclusions.} %% this section is mandatory
